---
title: "Worm genomes and life cycle transitions"
output: github_document
---

Check the phylogenies of exisiting helminth genomes to see how many life cycle length transitions occurred.

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(RColorBrewer)
library(ape)
library(phytools)

options(stringsAsFactors = FALSE)

# read data
dat <- read.csv(file = "../../data/combined_data_files_filledin.csv", header = TRUE, sep = ',')

# read tree
tree <- read.nexus(file = "platy_tree.nexus")
tree <- root(tree, outgroup = c("macrostomum.lignano", "schmidtea.mediterranea"), resolve.root = TRUE)
```


```{r}
# create tip state for mapping
dat <- mutate(dat, ps2 = sub(pattern = " ", replacement = ".", tolower(Parasite.species)))
mv <- match(tree$tip.label, dat$ps2)
tip.state <- dat$parasitism[mv]
names(tip.state)  <- dat$ps2[mv]

# drop tips with unmatched (e.g. species with two genomes)
tree <- drop.tip(tree, tip = tree$tip.label[is.na(mv)])
tip.state <- tip.state[!is.na(mv)]
```

Stochastic character mapping. Might not be the best approach...

```{r}
asr <- make.simmap(tree, x = tip.state)
```

Plot mapped states.

```{r}
mypalette <- brewer.pal(n = n_distinct(tip.state), "Set1")
names(mypalette) <- unique(tip.state)

plotSimmap(asr, colors = mypalette, pts = T, ftype = "off")
```

Counted transitions.

```{r}
countSimmap(asr)
```

Now try mapping life cycle complexity variable.

```{r}
dat <- mutate(dat, lcl.fac = if_else(lcl > 3, "3", as.character(lcl)))%>%
  mutate(lcl.fac = factor(lcl.fac, labels = c("free-living", "1-host", "2-host", "3-host")))
```

```{r}
tip.state <- dat$lcl.fac[mv]
names(tip.state)  <- dat$ps2[mv]
tip.state <- tip.state[!is.na(mv)]

test_trans_mat <- matrix(c(0,1,0,0,3,0,1,0,0,3,0,1,0,0,3,0), 4) # transitions to complexity more likely than reversals
asr <- make.simmap(tree, x = tip.state, 
                   pi = c(1, 0, 0, 0),
                   model = test_trans_mat)
```

```{r}
mypalette <- brewer.pal(n = n_distinct(tip.state), "Set1")
names(mypalette) <- unique(tip.state)

plotSimmap(asr, colors = mypalette, ftype = "off")
add.simmap.legend(colors = mypalette, vertical = F, prompt = F, leg = unique(tip.state))
```

Too much. Don't like this stochastic mapping.

```{r}
countSimmap(asr)
```

Add short branch to root.

```{r}
tree$edge.length <- if_else(tree$edge.length == 0, 0.001, tree$edge.length)
```


Parsimony reconstruction

```{r}
asr_mp <- MPR(tip.state, unroot(tree), outgroup = "macrostomum.lignano")
asr_mp
```

Likelihood reconstruction

```{r}
asr_lk <- ace(tip.state, tree, type = 'discrete', model = "ER")
asr_lk
```

Plot nice looking tree

```{r message=FALSE, warning=FALSE}
library(ggtree)
```

```{r}
p <- ggtree(tree) + 
  geom_tree() +
  geom_tiplab() 
p
```

Make species names look nicer.

```{r}
tree_df <- fortify(tree) # make a df

# nicer species names
n1 <- toupper(substr(tree_df$label, start = 1, stop = 1))
n2 <- substr(tree_df$label, start = 2, stop = nchar(tree_df$label))
nic_sp_names <- paste(n1, n2, sep = "")
nic_sp_names <- gsub(nic_sp_names, pattern = "\\.", replacement = " ")
nic_sp_names[which(nic_sp_names == "NANA")] <- NA
tree_df$label <- nic_sp_names
```

Plot tree with reconstructed ancestral states, starting with the likelihood reconstructions

```{r}
# ancestral state reconstructions to add to tree; branch colors
cols <- apply(asr_lk$lik.anc, 1, function(x) which(x == max(x)))
cols <- colnames(asr_lk$lik.anc)[cols]

# combine tip and branch states, add them to tree df (correct order)
state <- c(as.character(tip.state), cols)
tree_df$state <- state
tree_df <- mutate(tree_df, 
                  state = factor(state, levels = c("free-living", "1-host", "2-host", "3-host")))

# plot tree
p <- ggtree(tree_df) + 
  geom_tree(aes(color = state)) + 
  geom_tiplab(aes(color = state), size = 3) +
  xlim(0, max(tree_df$x) + 0.09) +
  scale_color_manual(values = mypalette) +
  theme(legend.position = c(0.25, 1), 
        legend.justification = c(1,1),
        legend.title = element_blank(),
        legend.text = element_text(size = 14))

p + geom_text2(aes(subset=!isTip, label=node), hjust=-.3)
```

Some of the branches near the root seem off. This is due to over-sampling of CLC parasites. But we know there are lots of 1-host cycle monogeneans between free-living and CLC parasites, so we can manually change these branches.

```{r}
tree_df[33:34,"state"] <- "1-host"
p <- ggtree(tree_df) + 
  geom_tree(aes(color = state)) + 
  geom_tiplab(aes(color = state), size = 3) +
  xlim(0, max(tree_df$x) + 0.09) +
  scale_color_manual(values = mypalette) +
  theme(legend.position = c(0.25, 1), 
        legend.justification = c(1,1),
        legend.title = element_blank(),
        legend.text = element_text(size = 14))
p
```

Can plot the parsimony reconstruction too.

```{r}
# ancestral states; MPR returns a matrix of states for parent-child nodes
# for coloring branches need to take value for root 'parent' and combine it with all the child values
cols <- c(asr_mp[1,1], asr_mp[,2])
cols <- c("free-living", "1-host", "2-host", "3-host")[cols]

# combine tip and branch states, add them to tree df (correct order)
state <- c(as.character(tip.state), cols)
tree_df$state <- state
tree_df <- mutate(tree_df, 
                  state = factor(state, levels = c("free-living", "1-host", "2-host", "3-host")))

# plot tree
p <- ggtree(tree_df) + 
  geom_tree(aes(color = state)) + 
  geom_tiplab(aes(color = state), size = 3) +
  xlim(0, max(tree_df$x) + 0.09) +
  scale_color_manual(values = mypalette) +
  theme(legend.position = c(0.25, 1), 
        legend.justification = c(1,1),
        legend.title = element_blank(),
        legend.text = element_text(size = 14))
p
```

Unlike in the likelihood case, the parsimony algorithm correctly assigned branches near the root. To this tree we'll add clade labels and points marking the life cycle transitions.

```{r}
# add clade labels
p <- p + 
  xlim(0, max(tree_df$x) + 0.19) +
  geom_cladelabel(node = 36, label = "Cestoda", offset = 0.13, color = "darkgray") +
  geom_cladelabel(node = 49, label = "Trematoda", offset = 0.13, color = "darkgray")
p
```

```{r}
# add points marking transitions
p + geom_text2(aes(subset=!isTip, label=node), hjust=-.3)
trans_nodes <- c(34, 49, 51, 37) # get nodes of transition

# add to plotting df
tree_df <- mutate(tree_df, trans_node = if_else(node %in% trans_nodes, TRUE, FALSE))

# add points onto plot
p <- p +
  geom_nodepoint(data = filter(tree_df, trans_node), 
                 shape = 19, size = 4, color = "gray", alpha = 0.5) 
p
ggsave(filename = "../../figs/platy_outtree.svg", width = 7.5, height = 5)
# fix labels in inkscape
```

