---
title: "Worm genomes and life cycle transitions"
output: github_document
---

Check the phylogenies of exisiting helminth genomes to see how many life cycle length transitions occurred.

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(RColorBrewer)
library(rotl)
library(ape)
library(phytools)

options(stringsAsFactors = FALSE)

dat <- read.csv(file = "data/combined_data_files_filledin.csv", header = TRUE, sep = ',')
```

Get the ott_id from tnrs.

```{r}
resolved_names <- tnrs_match_names(dat$Parasite.species, context_name = "Animals")
```

Check if ott_id is in the open tree of life. Checks if name is for example incertae sedis. One species is considered to be incorrect.

```{r}
genomes_in_tree <- is_in_tree(ott_id(resolved_names))
genomes_in_tree[!(genomes_in_tree)]
```

```{r}
inspect(resolved_names, taxon_name = "Schmidtea mediterranea")
```

Remove this species and download the subtree from the open tree of life. (have to add it separately...)

```{r}
resolved_names_noSchmidtea <- filter(resolved_names, unique_name != "Schmidtea mediterranea")
genome_tree <- tol_induced_subtree(ott_ids = resolved_names_noSchmidtea$ott_id, label_format = "name")
```

Plot the tree.

```{r}
plot(genome_tree, cex = 0.5)
```

Many polytomies. Could probably get more resolved tree from sequence data. It is just for illustration for now. Let's get the tree in shape for mapping.

```{r}
# remove trouble point from data
dat_noSchmidtea <- filter(dat, Parasite.species != "Schmidtea mediterranea")

# remove _ from tip labels, so it will match species names in data table
genome_tree$tip.label <- gsub(pattern = "_", replacement =  " ", x = genome_tree$tip.label)

# create tip state for mapping
mv <- match(genome_tree$tip.label, dat_noSchmidtea$Parasite.species)
tip.state <- dat_noSchmidtea$parasitism[mv]
names(tip.state)  <- dat_noSchmidtea$Parasite.species[mv]

# prep tree for mapping
  # making dichotomous is not needed
#tree.di <- multi2di(genome_tree, random = TRUE)
#Nnode <- tree.di$Nnode
#tree.di$node.label <- (1:Nnode+(1+Nnode))

tree.di <- compute.brlen(genome_tree) # simple branch lengths
```

Stochastic character mapping. Might not be the best approach...

```{r}
asr <- make.simmap(tree.di, x = tip.state)
```

Plot mapped states.

```{r}
mypalette <- brewer.pal(n = n_distinct(tip.state), "Set1")
names(mypalette) <- unique(tip.state)

plotSimmap(asr, colors = mypalette, pts = T, ftype = "off")
```

Counted transitions.

```{r}
countSimmap(asr)
```

Now try mapping life cycle complexity variable.

```{r}
dat_noSchmidtea <- mutate(dat_noSchmidtea, lcl.fac = if_else(lcl > 3, "3", as.character(lcl)))%>%
  mutate(lcl.fac = factor(lcl.fac, labels = c("free-living", "simple", "2-host", "3-host")))
```

```{r}
tip.state <- dat_noSchmidtea$lcl.fac[mv]
names(tip.state)  <- dat_noSchmidtea$Parasite.species[mv]

test_trans_mat <- matrix(c(0,1,0,0,3,0,1,0,0,3,0,1,0,0,3,0), 4) # transitions to complexity more likely than reversals
asr <- make.simmap(tree.di, x = tip.state, 
                   pi = c(1, 0, 0, 0),
                   model = test_trans_mat)
```

```{r}
mypalette <- brewer.pal(n = n_distinct(tip.state), "Set1")
names(mypalette) <- unique(tip.state)

plotSimmap(asr, colors = mypalette, ftype = "off")
add.simmap.legend(colors = mypalette, vertical = F, prompt = F, leg = unique(tip.state))
```


```{r}
countSimmap(asr)
```

```{r}
tree.di <- multi2di(tree.di)
tree.di <- compute.brlen(tree.di)
asr <- ace(tip.state, tree.di, type = 'discrete')
asr
```

```{r}
cols <- apply(asr$lik.anc, 1, function(x) which(x == max(x)))
cols <- colnames(asr$lik.anc)[cols]
branch_col <- mypalette[match(cols, names(mypalette))]
plot(tree.di, show.tip.label = F)
nodelabels(bg = branch_col)
```


Nothing very satisfying looking. Let's separate nematodes and platyhelminths

```{r}
#add discrete var for mapping
dat <- mutate(dat, lcl.fac = if_else(lcl > 3, "3", as.character(lcl)))%>%
  mutate(lcl.fac = factor(lcl.fac, labels = c("free-living", "simple", "2-host", "3-host")))

nem <- filter(dat, grepl(Clade, pattern = "Clade ")) # all nems begin with 'Clade'
platy <- filter(dat, !grepl(Clade, pattern = "Clade ")) # rest are platyhelminthes
```

Can try re-resolving names with more specific context_name argument

```{r}
nem_names <- tnrs_match_names(nem$Parasite.species, context_name = "Nematodes")
platy_names <- tnrs_match_names(platy$Parasite.species, context_name = "Platyhelminthes")
```

```{r}
nem_in_tree <- is_in_tree(ott_id(nem_names))
nem_in_tree[!(nem_in_tree)] # all in tree
```

```{r}
platy_in_tree <- is_in_tree(ott_id(platy_names))
platy_in_tree[!(platy_in_tree)]
```

```{r}
rm(nem_in_tree, platy_in_tree)
platy <- filter(platy, Parasite.species != "Schmidtea mediterranea")
platy_names <- filter(platy_names, unique_name != "Schmidtea mediterranea")
```

Get subtrees

```{r}
nem_tree <- tol_induced_subtree(ott_ids = nem_names$ott_id, label_format = "name")
platy_tree <- tol_induced_subtree(ott_ids = platy_names$ott_id, label_format = "name")
```


```{r}
# remove _ from tip labels, so it will match species names in data table
nem_tree$tip.label <- gsub(pattern = "_", replacement =  " ", x = nem_tree$tip.label)
platy_tree$tip.label <- gsub(pattern = "_", replacement =  " ", x = platy_tree$tip.label)

# add branch lengths
nem_treedi <- compute.brlen(nem_tree) # simple branch lengths
platy_treedi <- compute.brlen(platy_tree) # simple branch lengths

# tip states for mapping
mv <- match(nem_treedi$tip.label, nem$Parasite.species)
nem.tip.state <- nem$lcl.fac[mv]
names(nem.tip.state)  <- nem$Parasite.species[mv]

mv <- match(platy_treedi$tip.label, platy$Parasite.species)
platy.tip.state <- platy$lcl.fac[mv]
names(platy.tip.state)  <- platy$Parasite.species[mv]
```


```{r}
nem.asr <- make.simmap(nem_treedi, x = nem.tip.state, pi = c(1,0,0,0))
```

Plot mapped states.

```{r}
mypalette <- brewer.pal(n = n_distinct(nem.tip.state), "Set1")
names(mypalette) <- c("free-living", "simple", "2-host", "3-host")
plotSimmap(nem.asr, colors = mypalette, pts = F, ftype = "off")
add.simmap.legend(colors = mypalette, vertical = F, prompt = F, leg = unique(tip.state))
```


```{r}
platy.asr <- make.simmap(platy_treedi, x = platy.tip.state, pi = c(1,0,0,0))
```



```{r}
mypalette <- brewer.pal(n = n_distinct(platy.tip.state), "Set1")
names(mypalette) <- c("free-living", "simple", "2-host", "3-host")

plotSimmap(platy.asr, colors = mypalette, pts = F)
add.simmap.legend(colors = mypalette, vertical = F, prompt = F, leg = unique(tip.state))
```
